Equity conceptualizations and standards within transportation literature: a scoping review

### 2023-10-22 Manuscript/Manuscript.qmd
Antonio completed first exhaustive set of edits. This is now going back to Anastasia.

Meanwhile, firmly grounded on planet Earth, the employees that Bezos thanked for his suborbital escapade were struggling with some very mundane problems of their own, and none as lofty as conflicting schedules that prevented them from flying in rockets [@griffinPerson2021]. According to reports, people employed directly or indirectly by Amazon for warehouse or delivery work had, for years, been treated to "inhumane" conditions [@fungEuropean2018; @scottAmazon2019; @greeneAlgorithm2021], and subjected to surveillance on the job, degrading schedules that drove drivers to urinate in bottles, crushing demands for productivity quotas that led injury, facing little or no job security, and treated less with gratitude than as disposable inputs to feed Amazon's earnings and consumers' demands [@tungAmazon2020; @BBCNews2021; @reeseTeamsters2022; @middletonRobot2023].

@@

The question of _fairness_, however, is not a simple one, and most of us would probably have been stumped to explain in a clear and precise way, beyond a disquieting intuition, just why the above picture was disturbing. _Justice_ is a political ideal based on the principle that individuals should be treated in a fair and equitable manner [@gosslingUrban2016, p. 2], giving and receiving whatever they are _due_ [@jaggarPhilosophicalChallengesGlobal2009, pp. 1-2]. The political [and contested; @vanoutriveHow2019] nature of the concept presents challenges that are only narrowly amenable to scientific inquiry. For starters, the notion that people are "due" something depends on the values of a society, as embodied in its systems and institutions [@karnerFrom2020]. Values,in turn, are not subject to natural laws, but rather are the result of intersubjective realities, which is to say illusions whose legitimacy derives from a collective will to believe. For example, justice would likely mean something very different to a person in a democratic society, than to another in a society where they owed their all—perhaps including their own time or bodily autonomy—to some collective illusion (e.g., the state, a monarch). As well, the meaning of "justice" would likely differ in yet another society where very few owned most, and most owned very little due to a different illusion (e.g., that wealth equates merit). In these two hypothetical cases, elucidating the meaning of "fair" would in all certainty be beyond the dreams of most, since fair would be what the state, the monarch, or the extremely wealthy said it was. In contrast, in democratic societies^[Paraphrasing, democracy is the worst of all collective illusions, except for all others.] individuals rights are valued above the whims of the few. In such a setting, the task of defining a "just" distribution of the burdens and benefits of things—from income, to roads, to space travel—quickly becomes muddled, encumbered even, by the necessity to pay attention to a multitude of voices, not all of them equally loud.

@@

The objective of this report is to scan the state of knowledge in terms of defining and operationalizing "fairness" in the transportation domain. Much research has been devoted to the issues of _measuring _ equity in transportation, including (among many others) @ramjerdiEquity2006, @delboscUsing2011, @welchMeasure2013, @martensMeasuring2019, and @pritchardNew2022. Further, there are multiple sources that discuss the conceptual and philosophical foundations of equity and fairness in transportation [e.g., @martensTransport2016; @pereiraDistributiveJusticeEquity2017a; @vanoutriveHow2019]. Finally, previous reviews of planning documents have investigated equity from narrowly scoped perspectives, such as accessibility [@boisjolyHowGetThere2017] or a particular mode of transportation [e.g., cycling; @doranPursuitCyclingEquity2021]. These inquiries are valuable to scholar, planning agencies, and decision-makers alike, and the present review will tread similar, but not completely overlapping ground. In our estimation, there remains a gap in the literature in terms of understanding how standards for equity are developed and implemented in the transportation domain. In contrast, we do know that adoption of equity concepts in planning practice has lagged developments in academic work [@pereiraTransportationEquity2021; @boisjolyHowGetThere2017; @doranPursuitCyclingEquity2021; @linovskiEquityTransportationPlanning2020; @litmanEvaluatingTransportationEquity2022].

@@

These questions boil down to the development and use of _standards_ for transportation justice. The term "standard" is used to mean "something set up and established by authority as a rule for the measure of quantity, weight, extent, value, or quality"^[https://www.merriam-webster.com/dictionary/standard]. How much pollution is allowed to be generated, and where, depends on who is affected, and how much health is valued overall, as well as by whom. For example, firms may or may not adopt lower standards for the emission of pollutants in poorer areas; it might be that poor people end up being relegated to areas that adopt lower standards [@gouldsonFirms2006]. Regardless of the cause, the result is the same: pollution tends to be worse were poorer people are [@delucaCode2012].

Supporting the creation of (more) just transportation systems involves understanding the production, distribution, and management of transportation benefits and costs; how they are distributed; and what values are implemented (and by whom) in the form of standards [@pereiraDistributiveJusticeEquity2017a; @shellerMobilityJusticePolitics2018; @pereiraTransportationEquity2021]. Thus, for this review we engage the literature with the following questions in mind: 1) what is our current understanding of the things that transportation systems do, for whom, and to whom; 2) what does the literature say about the distribution of the benefits and costs of transportation systems; 3) and what values are embodied in normative statements about said distribution. Ultimately, this review aims to collate the existing academic knowledge on the matter, and present it in a manner useful to support the development and implementation of standards for equity in transportation planning and policy. In this, we aim to provide relief even, or particularly, to planners in those places where legislation explicitly calls for justice^[[As recently as -@martensFair2021, Martens and Golub note that Federal directives related to Title VI in the US "do not provide guidelines that can help agencies develop explicit standards to assess the distribution of accessibility benefits from projects or plans."]].

@@

After this introduction, we set the stage for our investigation by laying out some important definitions. We then describe the methods used for searching, selecting, and reviewing the relevant literature. 
<!--
In Section 2, we set the stage by defining what do we mean by equity dimensions, conceptualizations and types of standards used in the literature search and throughout the paper.
In Section 3, we outline the methods used in the scoping review conducted.
In Section 4, we summarize the findings of the literature.
In Section 5, we discuss the methods connects to standards, conceptualizations and dimensions.
In Section 6, we summarize the findings and link them to future transport planning agendas in addition to visioning ways the findings can be used to build *just* transportation planning processes.
-->

@@

Let us begin by stating that justice is an end goal, that is, a desirable state of affairs that we strive to achieve.

@@

It might be argued that all of the above are particular forms of distributive justice. Retributive justice, for example, is often concerned with the distribution of the benefits and burdens of being a member of society: the way it is usually achieved is by distributing intangibles like "freedom" (e.g., of movement, of association) as benefits, and/or claiming tangibles like money as burdens (e.g., fines). Reparative justice often distributes current benefits and burdens to redress past wrongs, for example by asking those who have benefited from past wrongs, even if unwittingly, to shoulder a bigger fiscal burden in order to cover programs that mete benefits to those who are still harmed by past wrongs. Procedural justice is the distribution of the benefits (e.g., the right to voice an opinion) and burdens (e.g., the effort required to develop an educated opinion) of the processes that lead to decisions with collective consequences.

We can then speak of the _purposes_ of distributive justice: to mete out retribution _fairly_, to repair past harms, and to ensure that procedures offer _equitable_ opportunities to influence outcomes. Equity and fairness are the instruments of justice, the tools by which society advances towards the end goal of justice.

@@

Fairness, in contrast to equity, is somewhat more complicated to define. The concept does not have the same history of development as an analytical tool, and can be interpreted in numerous, and possibly discordant ways. That this is the case is convincingly demonstrated by @martensFair2021 in their study of the application of Title VI of the Civil Rights Act of 1964 in accessibility planning practice in the US. Title VI explicitly talks about the distribution of benefits derived from Federal funding: "[n]o person in the United States shall, on the ground of race, color, or national origin, be excluded from participation in, be denied the benefits of, or be subjected to discrimination under any program or activity receiving Federal financial assistance." However, as @martensFair2021 show, there are several ways to comply with regulations while achieving different outcomes, ranging from the banal (do not _knowingly_ discriminate), to the substantive (compensation for past discrimination, i.e., reparative justice). What kind of justice does fairness serve in each case? It depends on what was the reason for seeking justice in the first place. Our reading of @martensFair2021 is that fairness is a yardstick that is best deployed _a priori_ than _a posteriori_, for doing the latter risks rationalizing the outcomes rather than driving them.

The last concept that we discuss in this section is that o a standard. Briefly, standards are a way of making concrete statements about fairness. Returning to the ambiguities in Title VI discussed by @martensFair2021, the attainment of justice depends on the standard used to indicate fairness. For example, explicit non-discrimination constitutes a very weak standard that takes aim at the actions of agencies instead of the recipients of the benefits; accordingly, any distribution of benefits would be considered fair, as long as the agency does not explicitly and knowingly targets or denies the benefits to particular groups. The standard provides conditions to determine whether a situation is _fair_. A similarly weak standard is a _Pareto improvement_, whereby it is possible to concentrate the benefits as long as no group is worse off compared to the status quo. A policy is fair as long it does no harm. A somewhat more strict standard, a Pareto-Plus improvement, stipulates that an intervention is fair when all groups receive at least _some_ (non-trivial) benefits; the size of the benefits for each group is irrelevant. In contrast to the notion of "do no harm", such a standard embodies the ideal that no one is denied benefits. An egalitarian standard would weigh the benefits or burdens by population, and fairness is achieved when each group give or receive in proportion to their size. An affirmative action standard is even stricter, since it requires the benefits to be distributed in a non-egalitarian way that favors those who are still harmed by past or present discriminatory practices.

@@

- **"Where?"** Traditionally related to the applicable domain or sphere of life relevant for justice. Conventionally this meant the in-group [e.g., members of the same nation state; @jaggarPhilosophicalChallengesGlobal2009, p. 3], a notion since expanded (e.g., with the Universal Declaration of Human Rights) to include all members of the human species, but has yet to cover other living creatures. <!-- SOUTH AMERICAN CONSTITUON THAT DOES INCLUDE IT NOTE FOR ANASTASIA--> In the case of justice in transportation, the question of "where?" is paramount; it might be argued that, by their very nature, transportation generates inequalities. By concentrating the effects of space-time convergence (for instance, by providing access to a transit system or a highway), an inequality is automatically generated. The burdens of transportation, in contrast, are often diffuse. They are incrementally paid, for example by a distributed population in the form of taxes by a spatially distributed population, or in poor health by people who were not party to transportation decisions made elsewhere.

@@

- When we ask **"Who?"**, we inquire about which entities should be regarded as subjects of justice <!-- OF INEQUITIES / MEASURING THE INEQUITIES OF 'WHO'. LIKE INCOME, THE MOBILITY TOOL IS MODIFIABLE -- THE MOBILITY TOOL IS NOT MODIFIABLE IN MANY CASES, IT IS WHO YOU ARE . T IS ALMOST LIKE A CHARACTERISTICS OF THE PERSON. PART OF THE SOCIAL CONTRACT THAT IDENTIFIES EOPEL WITH THEIR MOBILITY TOOL-->, meaning who are entitled to make claims of moral consideration from the perspective of justice. To make it tractable, this question is often approached through the filter of population groups, which may include the several traits, including gender identity, ableness, ethnicity, age, caste, and income. Often, it is appropriate to reflect on the intersections between traits, given evidence that the lived experiences of, say, a White woman and a Black women, can be markedly different. A possible complication in the case of transportation is that disentangling the "who" from the tools available to them is not always straightforward. Clearly, a person is not their mode of transportation; in practice, though, there are large segments of the population who live in situations where they cannot extricate themselves from the tools for mobility that they can use, either because they have driven themselves out of choices [see @laveryDriving2013], or have been driven out of choices by factors beyond their control, effectively ending up as captive users of a single mode [@jacquesRescuing2012; @cherancheryInvestigatingPerceptionCaptive2018]. In societies that have grown into transportation monocultures with a predilection for automobility [@millerCollaborative2011] there may actually be less choice about mobility tools than we would like to assume. So, while it is important to avoid conflation of the "who" with the "what",  for analytical purposes we need to be mindful of the connection between person and their mobility tools. In the case of transportatin, in addition to members of the public who use transportation systems, there is another category of **who**, that stand possibly in opposition to users, namely the entities charged with providing services, maintaining infrastructure, and so on. These could be ministries or departments of transportation, transit agencies, public works departments and others. Identifying these entities is relevant to ellucidate who is responsible for apportioning the benefits or mitigating the burdens of transportation.

- **"What?"** asks which entities should be regarded as objects of justice, meaning which kinds or categories of things should be distributed in a just manner. To understand the distributional implications of transportation systems, it is essential that we are clear about what they do. In other words, what do transportation systems _produce_? At their most fundamental, transportation systems are space-time convergence tools, tools that improve the rate at which time is traded for space. They usually do this by increasing the speed of movement: sidewalks facilitate walking, traffic lights facilitate the ordered flow of vehicular traffic, and a launching pad makes it possible for a rocket to takeoff. With complex interlocking parts (sidewalk, road, traffic sign, parking regulations), transportation systems produce _mobility_, the potential for movement.The realization of this potentials happens through travel but, as the adage goes, travel is derived demand, which seems to hold for most (even if not all) situations [e.g., @mokhtarianUnderstanding2001; @redmondPositive2001; @paezEnjoyment2010; @whalenMode2013]. For this reason, we cannot stop at considering mobility, but the ultimate goal of mobility, which is to reach destinations. In combination with land use systems (the spatial distribution of opportunities on the landscape), mobility produces _accessibility_, the potential to reach destinations. For this reason, we can think of the objects of transportation justice as being _proximate_ (the tools of mobility, mobility itself), and _ulterior_ (accessibility, opportunities for activity participation). The burdens of transportation are also many and varied. Some are direct and paid directly by the traveler (e.g., travel time, out-of-pocket costs), but many others are indirect and related to network externalities (e.g., exposure to pollution).

- The next question is **"How?"**, and it relates to the allocation of various objects of justice ("what") to various subjects of justice ("who") in various circumstances. **Equity standards** are a tool for answering this distributive question: how do we allocate burdens and benefits and to whom? Standards are thresholds that when operationalized effectively define what is in/equitable. The thresholds can be quantitative (e.g., square meters of green space per capita), or they can be qualitative descriptions (e.g., do not knowingly discriminate), or a mixed of the two. Some examples include: maximum travel distance/cost/time to or from key destinations, levels of maximum exposure to externalities (i.e., noise or air pollution),un/fulfilled needs, and dis/satisfaction with travel. A number of theoretical and conceptual frameworks exist to support us when approaching this question, and we can draw from concepts in transport-related social exclusion, transport disadvantage, and/or transport poverty, which are typically based on equity principles, such as utilitarianism, Sen's capabilities approach, and sufficientarism.

- Lastly, convincing answers to the above questions require a supporting rationale: a **"Why?"** [@jaggarPhilosophicalChallengesGlobal2009]. This is perhaps the most slippery of all the questions posed here. Justice is an inherently social construct <!--, and does not exist outside of human society-->. Asking **why?** amounts to asking what sort of social contracts regulate human interactions. These contracts can be defined by constitution, but there are often unwritten and possibly contested variants. To give an example drawing from the Canadian Bill of Rights, a number of rights and fundamental freedoms [including the liberty to move freely; see @chartopediaArt72023] are recognized to exist in Canada "without discrimination by reason of race, national origin, colour, religion or sex". Notice that this declaration includes several individual traits that define the subjects of justice, but does not consider age as one aspect of the person. Does this mean that they apply universally irrespective of age? Surely not, since no reasonable person would consider a toddler's demands for freedom of association as an absolute. So, when do these rights fully apply in the life of a person? In 2017, Adrian Crook of Vancouver, B.C., was warned by the province's Ministry of Children and Family Development that his kids could not be out of home in the community, alone or with other kids the same age, without supervision. The Ministry's argument was that children riding the bus unsupervised compromised the parent's ability to provide care and placed them at risk [@brendHere2017]. Crook argued that the goal in teaching his kids (then aged 7, 8, 9 and 11) to use transit to go to school was to raise self-reliant children [@stueckYoung2017]. While both sides discovered that there were not clear rules about kids riding transit alone [@brendHere2017], statistically there is evidence that travel by car is riskier than travel by bus [@morencyTraveling2018]. If the public good is to reduce risk, should not children be banned from riding in cars until they are ten? In a functioning democracy like Canada, we have chosen to care about fundamental freedoms and rights. This provides an answer as to "why" we would even consider children as subjects of justice. But the devil is in the details, and the other questions discussed above are needed to pin the devil by the tail. Are children under the age of 10 subjects of justice (i.e., are they a "who")? If so, is the ability to use transit unsupervised an object of justice (i.e., the "what")? When and where do these questions apply (e.g., riding the bus to school, or to a social event; during the day, or late at night)? Eventually, in 2020, a court of appeals ruled in Crook's favor, allowing his children to continue to ride the bus, thus providing much needed clarity with respect to the use of transit by children who desire independent mobility [@stueckCourt2020]. In actuality, the courts provided a standard of equity: starting at a certain age, children are objects of justice from the perspective of freedom of movement, and they are due the benefits of independent mobility by transit.

@@

This review examines the breadth and depth of the academic literature on transportation to identify the extent to which standards for equity are defined and employed. In this task, we follow the Joanna Briggs Institute (JBI) approach to the conduct of scoping reviews. This approach builds upon the Arksey and O'Malley (2005) framework [@petersUpdatedMethodologicalGuidance2020]. The JBI is .

Further, the review is guided too by the Preferred Reporting Items for Systematic Reviews and Meta-Analyses extension for scoping reviews (PRISMA-ScR), which is consistent with the JBI approach [@triccoPRISMAExtensionScoping2018]. The use of these methods allows us to explore, in a consistent and organized manner, a relatively specialized topic within the broader transportation literature. In this way, we aim to collate the current knowledge as found in the landscape of published research.

The primary research question and the protocol were initially defined in consultation among the authors of the report. In other words, the starting point was the level of knowledge of experts in the subject matter. The initial draft of the protocol was refined from preliminary searches of related-reviews [e.g., @iglesiasHowUnevenUrban2019; @sagarisUsingPARFrame2020; @vecchioAccessibilityCapabilitiesApproach2021], and in consultation with a University of Toronto Research Services Librarian and Liaison Librarian in City Studies.

@@

<!--

EDITED UP TO THIS POINT.

@@

We discuss trends in how the literature has conceptualized and used standards (the "How") to evaluate the dimensions (the "What") of transportation equity. A noteable point is the trends are context-specific: they concern the "When" and "Where" (differences across publication year, across case studies continent, and between urban and rural contexts) and the "Who" (differences in the focus on the population sub-groups, transport modes, destinations of interest) as discussed in the preceding sections following the four equity dimensions that structure this review (mobility/accessibility, traffic-related pollution, human-health related, and traffic safety).

>>>>

Meanwhile, firmly grounded on planet Earth, the employees that Bezos thanked for his suborbital escapade were struggling with some very mundane problems of their own, and none as lofty as conflicting schedules that prevented them from flying in rockets [@griffinPerson2021]. According to reports, people employed directly or indirectly by Amazon for warehouse or delivery work had, for years, been treated to "inhumane" conditions [@fungEuropean2018; @scottAmazon2019; @greeneAlgorithm2021], and subjected to surveillance on the job, degrading schedules that drove drivers to urinate in bottles, crushing demands for productivity quotas that led to injury, all the while facing little or no job security. From this perspective, they were treated less with gratitude and more as disposable inputs to feed Amazon's earnings and consumers' demands [@tungAmazon2020; @BBCNews2021; @reeseTeamsters2022; @middletonRobot2023].

@@

The question of _fairness_, however, is not a simple one, and most of us would probably have been stumped to explain in a clear and precise way, beyond a disquieting intuition, just why the above picture was disturbing. _Justice_ is a political ideal based on the principle that individuals should be treated in a _fair_ and _equitable_ manner [@gosslingUrban2016, p. 2], giving and receiving whatever they are _due_ [@jaggarPhilosophicalChallengesGlobal2009, pp. 1-2]. The political [and contested; @vanoutriveHow2019] nature of the concept presents challenges that are only narrowly amenable to scientific inquiry. For starters, the notion that people are "due" something depends on the values of a society, as embodied in its systems and institutions [@karnerFrom2020]. Values, in turn, are not subject to natural laws, but rather are the result of intersubjective realities, which is to say illusions whose legitimacy derives from a collective will to believe. For example, justice would likely mean something very different to a person in a democratic society, than to another in a society where they owed their all—perhaps including their own time or bodily autonomy—to some collective illusion (e.g., the state, or a monarch). As well, the meaning of "justice" would likely differ in yet another society where very few owned most, and most owned very little due to a different illusion (e.g., that wealth equates merit). In these two hypothetical cases, elucidating the meaning of "fair" would in all certainty be beyond the dreams of most, since fair would be what the state, the monarch, or the extremely wealthy said it was. In contrast, in democratic societies^[Paraphrasing, democracy is the worst of collective illusions, except for all others.] individuals rights are valued above the whims of the few. In such a setting, the task of defining a "just" distribution of the burdens and benefits of things—from income, to roads, to space travel—quickly becomes muddled, encumbered even, by the necessity to pay attention to a multitude of voices, not all of them equally loud.

@@

The objective of this report is to scan the state of knowledge in terms of defining and operationalizing "fairness" in the transportation domain. Much research has been devoted to the issues of _measuring _ equity in transportation, including (among many others) @ramjerdiEquity2006, @delboscUsing2011, @welchMeasure2013, @martensMeasuring2019, and @pritchardNew2022. Further, there are multiple sources that discuss the conceptual and philosophical foundations of equity and fairness in transportation [e.g., @martensTransport2016; @pereiraDistributiveJusticeEquity2017a; @vanoutriveHow2019]. Finally, previous reviews of planning documents have investigated equity from narrowly scoped perspectives, such as accessibility [@boisjolyHowGetThere2017] or a particular mode of transportation [e.g., cycling; @doranPursuitCyclingEquity2021]. These inquiries are valuable to scholars, planning agencies, the public, and decision-makers alike, and the present review will tread similar, but not completely overlapping ground. In our estimation, there remains a gap in the literature in terms of understanding how standards for equity are developed and implemented in the transportation domain. In contrast, we do know that adoption of equity concepts in planning practice has lagged developments in academic work [@pereiraTransportationEquity2021; @boisjolyHowGetThere2017; @doranPursuitCyclingEquity2021; @linovskiEquityTransportationPlanning2020; @litmanEvaluatingTransportationEquity2022].

@@

These questions boil down to the development and use of _standards_ for transportation justice. The term "standard" connotes "something set up and established by authority as a rule for the measure of quantity, weight, extent, value, or quality"^[https://www.merriam-webster.com/dictionary/standard]. How much pollution is allowed to be generated, and where, depends on who is affected, and how much health is valued overall, as well as by whom. For example, firms may or may not adopt lower standards for the emission of pollutants in poorer areas; it might be that poor people end up being relegated to areas that already had lower standards [@gouldsonFirms2006]. Regardless of the cause, the result is the same: pollution tends to be worse were poorer people are [@delucaCode2012].

Supporting the creation of (more) just transportation systems involves understanding the production, distribution, and management of transportation benefits and costs; how they are distributed; and what values are implemented (and by whom) in the form of standards [@pereiraDistributiveJusticeEquity2017a; @shellerMobilityJusticePolitics2018; @pereiraTransportationEquity2021]. Thus, for this review we engage the literature with the following questions in mind: 1) what is our current understanding of the things that transportation systems do, for whom, and to whom; 2) what does the literature say about the distribution of the benefits and costs of transportation systems; 3) and what values are embodied in normative statements about said distribution. Ultimately, this review aims to collate the existing academic knowledge on the matter, and present it in a manner useful to support the development and implementation of standards for equity in transportation planning and policy. In this, we aim to provide relief even, or particularly, to planners in those places where calls for justice are explicitly made through legislation^[[As recently as -@martensFair2021, Martens and Golub note that Federal directives related to Title VI in the US "do not provide guidelines that can help agencies develop explicit standards to assess the distribution of accessibility benefits from projects or plans."]].

@@

After this introduction, we set the stage for our investigation by laying out some important definitions. We then describe the methods used for searching, selecting, and reviewing the relevant literature. This is followed by a description of the landscape surveyed, which is then appraised critically. We will conclude with some calls for action to improve the practice of setting and using standards of fairnes for transportation justice.

@@

Let us begin by stating that justice is an end goal, that is, a desirable state of affairs that we are morally obligated to achieve.

@@

It might be argued that all of the above are particular forms of distributive justice. Retributive justice, for example, is often concerned with the distribution of the benefits and burdens of being a member of society; the way it is usually achieved is by distributing intangibles like "freedom" (e.g., of movement, of association) as benefits, and/or claiming tangibles like resources as burdens (e.g., fines). Reparative justice often distributes current benefits and burdens to redress past wrongs, for example by asking those who have benefited from said wrongs, even if unwittingly, to shoulder a bigger fiscal burden in order to cover programs that mete benefits to those who are still harmed by past wrongs. Procedural justice is the distribution of the benefits (e.g., the right to voice an opinion) and burdens (e.g., the effort required to develop an educated opinion) of the processes that lead to decisions with collective consequences.

We can then speak of the _purposes_ of distributive justice: to mete out retribution _fairly_, to repair past harms, and to ensure that procedures offer _equitable_ opportunities to influence outcomes. Equity and fairness from this perspective are the instruments of justice, the tools by which society advances towards the end goal of justice.

@@

Fairness, in contrast to equity, is somewhat more complicated to define. The concept does not have the same history of development as an analytical tool, and can be interpreted in numerous, and possibly discordant ways. That this is the case is convincingly demonstrated by @martensFair2021 in their study of the application of Title VI of the Civil Rights Act of 1964 in accessibility planning in the US. Title VI explicitly talks about the distribution of benefits derived from Federal funding: "[n]o person in the United States shall, on the ground of race, color, or national origin, be excluded from participation in, be denied the benefits of, or be subjected to discrimination under any program or activity receiving Federal financial assistance." However, as @martensFair2021 show, there are several ways to comply with regulations while achieving different outcomes, ranging from the banal (do not _knowingly_ discriminate), to the substantive (compensation for past discrimination, i.e., reparative justice). What kind of justice does fairness serve in each case? It depends on what was the reason for seeking justice in the first place. Our reading of @martensFair2021 is that fairness is a yardstick that is best deployed _a priori_ than _a posteriori_, for doing the latter risks rationalizing the outcomes rather than driving them.

The last concept that we discuss in this section is that of a standard. Briefly, standards are a way of making concrete statements about fairness. Returning to the ambiguities in Title VI discussed by @martensFair2021, the attainment of justice depends on the standard used to indicate fairness. For example, explicit non-discrimination constitutes a very weak standard that takes aim at the actions of agencies instead of the recipients of the benefits; accordingly, any distribution of benefits would be considered fair, as long as the agency does not explicitly and knowingly target or deny benefits to particular groups. The standard provides conditions to determine whether a situation is _fair_. A similarly weak standard is a _Pareto improvement_, whereby it is possible to concentrate the benefits as long as no group is worse off compared to the status quo. A policy is fair as long it does no harm. A somewhat more strict standard, a Pareto-Plus improvement, stipulates that an intervention is fair when all groups receive at least _some_ (non-trivial) benefits; the size of the benefits for each group is irrelevant. In contrast to the notion of "do no harm", such a standard embodies the ideal that no one is denied benefits. An egalitarian standard would weigh the benefits or burdens by population, and fairness is achieved when each group give or receive in proportion to their size. In contrast, an affirmative action standard is even stricter, since it requires the benefits to be distributed in a non-egalitarian way that favors those who are still harmed by past or present discriminatory practices.

@@

- **"Where?"** Traditionally related to the applicable domain or sphere of life relevant for justice. Conventionally this meant the in-group [e.g., members of the same nation state; @jaggarPhilosophicalChallengesGlobal2009, p. 3], a notion since expanded (e.g., with the Universal Declaration of Human Rights) to include all members of the human species. In the case of justice in transportation, the question of "where?" is paramount; it might be argued that, by their very nature, transportation generates inequalities. By concentrating the effects of space-time convergence (for instance, by providing access to a transit system or a highway), an inequality is automatically generated. The burdens of transportation, in contrast, are often diffuse. They are incrementally paid, for example by a distributed population in the form of taxes, or by a population with a different spatial distribution in the form of poor health.

@@

- When we ask **"Who?"**, we inquire about which entities should be regarded as subjects of justice,<!-- OF INEQUITIES / MEASURING THE INEQUITIES OF 'WHO'. LIKE INCOME, THE MOBILITY TOOL IS MODIFIABLE -- THE MOBILITY TOOL IS NOT MODIFIABLE IN MANY CASES, IT IS WHO YOU ARE . T IS ALMOST LIKE A CHARACTERISTICS OF THE PERSON. PART OF THE SOCIAL CONTRACT THAT IDENTIFIES EOPEL WITH THEIR MOBILITY TOOL--> meaning those entitled to make claims of moral consideration from the perspective of justice. To make it tractable, this question is often approached through the filter of population groups, which may include several concurrent traits, such as gender identity, ableness, ethnicity, age, caste, and income. Often, it is appropriate to reflect on the intersections between traits, given evidence that the lived experiences of, say, a White woman and a Black women, can be markedly different between each other, in addition to being different from those of men. A possible complication in the case of transportation is that disentangling the "who" from the tools available to them is not always straightforward. Clearly, a person is not their mode of transportation; in practice, though, there are large segments of the population who live in situations where they cannot extricate themselves from the tools for mobility that they can use, either because they have driven themselves out of choices [see @laveryDriving2013], or have been driven out of choices by factors beyond their control, effectively ending up as captive users of a single mode [@jacquesRescuing2012; @cherancheryInvestigatingPerceptionCaptive2018]. In societies that have grown into transportation monocultures with a predilection for automobility [@millerCollaborative2011] there may actually be less choice about mobility tools than we would like to assume. So, while it is important to avoid conflating the "who" with the "what", for analytical purposes we need to be mindful of the connection between person and their mobility tools. In the case of transportation, in addition to members of the public who use transportation systems, there is another category of **who**, that stands possibly in opposition to users, namely the entities charged with providing services, maintaining infrastructure, and so on. These could be ministries or departments of transportation, transit agencies, public works departments and others. Identifying these entities is relevant to ellucidate who is responsible, for example for apportioning the benefits or mitigating the burdens of transportation.

- **"What?"** asks which entities should be regarded as objects of justice, meaning which kinds or categories of things should be distributed in a just manner. To understand the distributional implications of transportation systems, it is essential that we are clear about what they do. In other words, what do transportation systems _produce_? At their most fundamental, transportation systems are space-time convergence technologies, tools that improve the rate at which time is traded for space. They usually do this by increasing the speed of movement: sidewalks facilitate walking, traffic lights facilitate the ordered flow of vehicular traffic, and a launching pad makes it possible for a rocket to take off. With complex interlocking parts (sidewalk, road, traffic sign, parking regulations), transportation systems produce _mobility_, the potential for movement. The realization of this potentials happens through travel. However, as the adage goes, travel is derived demand, which seems to hold for most (even if not all) situations [e.g., @mokhtarianUnderstanding2001; @redmondPositive2001; @paezEnjoyment2010; @whalenMode2013]. For this reason, we cannot stop at considering mobility, but the ulterior goal of mobility, which is to reach destinations. In combination with land use systems (the spatial distribution of opportunities on the landscape), mobility produces _accessibility_, the potential to reach destinations. In this manner, we can think of the objects of transportation justice as being _proximate_ (the tools of mobility, mobility itself), and _ulterior_ (accessibility, opportunities for activity participation). The burdens of transportation are also many and varied. Some are direct and paid directly by the traveler (e.g., travel time, out-of-pocket costs), but many others are indirect and related to network externalities (e.g., exposure to pollution).

- The next question is **"How?"**, and it relates to the allocation of various objects of justice ("what") to various subjects of justice ("who") in various circumstances ("when" and "where"). **Equity standards** are a tool for answering this distributive question: how do we allocate burdens and benefits and to whom? Standards are thresholds that when operationalized effectively define what is fair, that is, in/equitable. The thresholds can be quantitative (e.g., square meters of green space per capita), or they can be qualitative descriptions (e.g., do not knowingly discriminate), or a mix of the two. Some examples include: maximum travel distance/cost/time to or from key destinations, levels of maximum exposure to externalities (i.e., noise or air pollution),un/fulfilled needs, and dis/satisfaction with travel. A number of theoretical and conceptual frameworks exist to support us when approaching this question, and we can draw from concepts in transport-related social exclusion, transport disadvantage, and/or transport poverty, which are typically based on equity principles, such as utilitarianism, Sen's capabilities approach, and sufficientarism.

- Lastly, convincing answers to the above questions require a supporting rationale: a **"Why?"** [@jaggarPhilosophicalChallengesGlobal2009]. This is perhaps the most slippery of all the questions posed here. Justice is an inherently social construct <!--, and does not exist outside of human society-->. Asking **why?** amounts to asking what sort of social contracts regulate human interactions, that is, what are the rules that our collective will to believe impose on each of us. These contracts can be defined by constitution, but there are often unwritten and possibly contested variants. To give an example drawing from the Canadian Bill of Rights, a number of rights and fundamental freedoms [including the liberty to move freely; see @chartopediaArt72023] are recognized to exist in Canada "without discrimination by reason of race, national origin, colour, religion or sex". Notice that this declaration includes several individual traits that define the subjects of justice, but does not consider age as one aspect of the person. Does this mean that they apply universally irrespective of age? Surely not, since no reasonable person would consider a toddler's demands for freedom of association as an absolute. So, when do these rights fully apply in the life of a person? In 2017, Adrian Crook of Vancouver, B.C., was warned by the province's Ministry of Children and Family Development that his kids could not be out of home in the community, alone or with other kids the same age, without supervision. The Ministry's argument was that children riding the bus unsupervised compromised the parent's ability to provide care and placed them at risk [@brendHere2017]. Crook argued that the goal in teaching his kids (then aged 7, 8, 9 and 11) to use transit to go to school was to raise self-reliant children [@stueckYoung2017]. While both sides discovered that there no clear rules existed about kids riding transit alone [@brendHere2017], statistically there is evidence that travel by car is riskier than travel by bus [@morencyTraveling2018]. If the public good is to reduce risk, should not children be banned from riding in cars until they are ten? In a functioning democracy like Canada, people have chosen to care about fundamental freedoms and rights. This provides an answer as to "why" we would even consider children as subjects of justice. But the devil is in the details, and answering the other questions discussed above is essential to pin the devil by the tail. Are children under the age of 10 subjects of justice (i.e., are they a "who")? If so, is the ability to use transit unsupervised an object of justice (i.e., the "what")? When and where do these questions apply (e.g., to to to school, or to a social event; during the day, or late at night)? Eventually, in 2020, a court of appeals ruled in Crook's favor, allowing his children to continue to ride the bus, thus providing much needed clarity with respect to the use of transit by children who desire independent mobility [@stueckCourt2020]. In actuality, the courts _created_ a standard of equity: starting at a certain age, children are objects of justice from the perspective of freedom of movement, and they are due the benefits of independent mobility by transit.

@@

This review examines the breadth and depth of the academic literature on transportation to identify the extent to which standards for equity are defined and employed. In this task, we follow the Joanna Briggs Institute (JBI) approach to the conduct of scoping reviews, an approach that builds upon the Arksey and O'Malley (2005) framework [@petersUpdatedMethodologicalGuidance2020].

Further, the review is also guided by the Preferred Reporting Items for Systematic Reviews and Meta-Analyses, particularly the extension for scoping reviews (PRISMA-ScR), which is consistent with the JBI approach [@triccoPRISMAExtensionScoping2018]. The use of these methods allows us to explore, in a consistent and organized manner, a relatively specialized topic within the broader transportation literature. In this way, we aim to collate the current knowledge as found in the landscape of published research.

The primary research question and the protocol were initially defined in consultation among the authors of the report. In other words, the starting point for the review was the level of knowledge of a group of experts in the field of transportation. The initial draft of the protocol was refined from preliminary searches of related-reviews [e.g., @iglesiasHowUnevenUrban2019; @sagarisUsingPARFrame2020; @vecchioAccessibilityCapabilitiesApproach2021], and in consultation with a University of Toronto Research Services Librarian and a Liaison Librarian in City Studies.

@@

@@

Applying the analytical framework to questions of justice in transportation yields a number of insights. Here, we critically review some of the trends that we identified from the literature in terms of concepts and standards (the "how") and how they apply to the objects (the "what") and subjects (the "who") of justice, as well as under which situations ("when" and "where" ).
